Natural sources of vagueness and their implications

### 2017-04-05 book_chapter/paper.tex
Shortened intro to signaling games

Signaling games were first introduced as models of communication by David Lewis~\parencite*{lewis_convention_1969}.
His original objective was to address an argument raised by W.~V.~Quine~\parencite*{quine_truth_1936} and others against the possibility of language having started as a conventional system: if language is a convention, it had to be originally established by an agreement; in order to establish an agreement, a convention-governed system of communication would have to already have been in place; thus, although some languages could have been established by agreement if another convention was already in place, not all of them could.
To this, Lewis retorts:
\begin{quote}
I offer this rejoinder: an agreement sufficient to create a convention need not be a transaction involving language or any other conventional activity.
All it takes is an exchange of manifestations of a propensity to conform to a regularity.%
~\parencite*[87--88]{lewis_convention_1969}
\end{quote}
In order to support this claim, Lewis studies coordination problems formalized in terms of game theory.
These are ``situations of interdependent decision by two or more agents in which coincidence of interest predominates and in which there are two or more proper coordination equilibria''~\parencite*[24]{lewis_convention_1969}.
In game theory terms, the agents interested in the coordination are the players, the game involves each player making an independent choice from his set of available actions, a payoff is what is attributed to each player based on the choices of both.
Adapting one of Lewis' examples, say Alice and Bob want to get together.
They usually meet at either Caf\'e One or Bistro Two.
Imagine there is no way for them to make any explicit agreement about it.
They are thus left to independently decide to either go to Caf\'e One or go to Bistro Two and hope for the other to show up there.
Neither has any preference for either place, but they do want to meet.
Each thus prefers to go to one of the places only if the other also decides to go to that particular place.

The setup of a coordination problem, \emph{i.e.}~the available actions and the relative interests of the players, can be represented in a payoff matrix, where rows show one player's available actions, columns the other player's available actions, and each cell gives the payoff for each of the players based on its row and column combination.
For example, consider the following matrix:
\begin{center}
\begin{tabular}{r|c|c|}
\multicolumn{1}{r}{}
 & \multicolumn{1}{c}{$b_1$}
 & \multicolumn{1}{c}{$b_2$} \\ \cline{2-3}
   $a_1$ & $1,1$ & $0,0$ \\ \cline{2-3}
   $a_2$ & $0,0$ & $1,1$ \\ \cline{2-3}
\end{tabular}
\end{center}
We can see this as representing the following game: one player has actions $A = \lbrace a_1, a_2 \rbrace$ available, the other $B = \lbrace b_1, b_2 \rbrace$; they prefer to coordinate $a_1$ with $b_1$ or $a_2$ with $b_2$, thus if this is achieved each gets a payoff of $1$, otherwise they each get $0$.
Connecting with the example laid out above, think of $a_1$ as Alice going to Caf\'e One, $a_2$ as Alice going to Bistro Two, $b_1$ as Bob going to Caf\'e One, and finally $b_2$ as Bob going to Bistro Two; their payoff is $1$ for both if they coordinate on the place to meet (independently of which one), and $0$ if they miscoordinate.
Formally, all we need to define in order to characterize such a situation are the sets of actions $A$ and $B$, and a utility function $U : A \times B \rightarrow \mathbb{R}^2$ that specifies the payoff for each player given the choices of actions.

The above example is a very simple case, but it serves to illustrate Lewis' notion of convention.
The two pairs of choices $(a_1,b_1)$ and $(a_2,b_2)$ are stable coordination equilibria, because in such a scenario no player has an incentive to unilaterally change his choice: if the first player is going to choose action $a_1$, the second player would get a payoff of $0$ for switching to $b_2$, instead of $1$ from sticking to $b_1$; the same reasoning applies, mutatis mutandis, to the other player and the other equilibrium.
However, neither player has a preference between those two pairs.
If the same coordination problem arises again, Lewis says, we can expect precedence to induce a kind of regularity: if the players manage to coordinate on one of the two equilibria, they should be expected to repeat the choices of actions that lead to that success, thus remaining in the same equilibrium.
Not only can we expect the players in such a repeated interaction to conform to a regularity, but also to expect others to do the same.
Their preference is to remain in a certain equilibrium given that others do too.
This summarizes Lewis's notion of a convention.

In order to extend this notion to \emph{linguistic} conventions, Lewis considers situations where the actions available involve sending and receiving signals or messages.
Thus, we could think of two players with different roles.
The first player, which Lewis calls the communicator~\parencite*[130]{lewis_convention_1969} and we will here call the sender, has knowledge about which of a number of possible states of affairs obtains and, depending on this information, chooses a signal to send.
The second player, which Lewis calls the audience~\parencite*[130]{lewis_convention_1969} and we will here call the receiver, has knowledge about which signal the sender chose and, based on this information, chooses one of several possible responses.
A preference relation exists between responses and states of affairs, and a payoff is attributed to each player based on the choices of both.
Note that Lewis assumes that no player has any preference regarding the particular signal that is used, provided that it enables coordination.
Formally, in order to describe the setup all we need is to specify a set of possible states of affairs $T$, a set of available signals or messages $M$, a set of responses or actions $A$, and the utility function $U : T \times A \rightarrow \mathbb{R}^2$.

Despite the added dimension of the signal exchange, these so-called signaling problems can be seen as particular cases of coordination problems if we consider the players' choices to be of contingency plans or strategies.
A communicator's contingency plan, or sender strategy as we will call it, is a specification of a choice of message for each possible state of affairs.
It thus describes the sender's behavior conditional on the state of affairs that obtains.
An audience's contingency plan, or receiver strategy as we will call it, analogously specifies a choice of action for each possible message.
Thus, formally, what the sender chooses is a function $\sigma : T \rightarrow M$ and the receiver a function $\rho : M \rightarrow A$.
The expected utility $EU$ of a pair of strategies $(\sigma,\rho)$ can be calculated using the utility function between states of affairs and actions as a sum of the payoffs for all cases. %, \emph{i.e.}:
% $$
% EU(\sigma, \rho) = \sum_{t \in T} U(t, \rho(\sigma(t)))
% $$
As an example, consider a game with $T = \lbrace t_1, t_2 \rbrace$, $M = \lbrace m_1, m_2 \rbrace$, $A = \lbrace a_1, a_2 \rbrace$, and the following utility matrix:
\begin{center}
\begin{tabular}{r|c|c|}
\multicolumn{1}{r}{}
 & \multicolumn{1}{c}{$a_1$}
 & \multicolumn{1}{c}{$a_2$} \\ \cline{2-3}
   $t_1$ & $1,1$ & $0,0$ \\ \cline{2-3}
   $t_2$ & $0,0$ & $1,1$ \\ \cline{2-3}
\end{tabular}
\end{center}
Possible sender and receiver strategies are, for example, $\sigma = \lbrace t_1 \mapsto m_2, t_2 \mapsto m_1 \rbrace$ and $\rho = \lbrace m_1 \mapsto a_2, m_2 \mapsto a_1 \rbrace$.
These would have an expected utility of $2$ for both sender and receiver, since when $t_1$ obtains the sender will use $m_2$ and to this message the receiver will respond with $a_1$ which achieves a payoff of $1$, when $t_2$ obtains the sender will use $m_1$ and to this message the receiver will respond with $a_2$ which also achieves a payoff of $1$.
% the following expected utility:
% \begin{equation}
% \begin{split}
% EU(\sigma, \rho) & = \sum_{t \in T} U(t, \rho(\sigma(t))) \\
%                  & = U(t_1, \rho(\sigma(t_1))) + U(t_2, \rho(\sigma(t_2))) \\
%                  & = U(t_1, \rho(m_2)) + U(t_2, \rho(m_1)) \\
%                  & = U(t_1, t_1) + U(t_2, t_1) \\
%                  & = (1,1) + (1,1) \\
%                  & = (2,2)
% \end{split}
% \end{equation}
% Based on this, we could also create a matrix of expected utilities for this example as follows:
% \begin{center}
% \begin{tabular}{r|c|c|c|c|}
% \multicolumn{1}{r}{}
%  & \multicolumn{1}{c}{$m_1 \mapsto a_1$}
%  & \multicolumn{1}{c}{$m_1 \mapsto a_1$}
%  & \multicolumn{1}{c}{$m_1 \mapsto a_2$}
%  & \multicolumn{1}{c}{$m_1 \mapsto a_2$} \\
% \multicolumn{1}{r}{}
%  & \multicolumn{1}{c}{$m_2 \mapsto a_1$}
%  & \multicolumn{1}{c}{$m_2 \mapsto a_2$}
%  & \multicolumn{1}{c}{$m_2 \mapsto a_1$}
%  & \multicolumn{1}{c}{$m_2 \mapsto a_2$} \\ \cline{2-5}
%    $t_1 \mapsto m_1, t_2 \mapsto m_1$ & $1,1$ & $1,1$ & $1,1$ & $1,1$ \\ \cline{2-5}
%    $t_1 \mapsto m_1, t_2 \mapsto m_2$ & $1,1$ & $2,2$ & $0,0$ & $1,1$ \\ \cline{2-5}
%    $t_1 \mapsto m_2, t_2 \mapsto m_1$ & $1,1$ & $0,0$ & $2,2$ & $1,1$ \\ \cline{2-5}
%    $t_1 \mapsto m_2, t_2 \mapsto m_2$ & $1,1$ & $1,1$ & $1,1$ & $1,1$ \\ \cline{2-5}
% \end{tabular}
% \end{center}
They also represent one of the two stable conventions in this game, the other being the pair of strategies $\sigma = \lbrace t_1 \mapsto m_1, t_2 \mapsto m_2 \rbrace$ and $\rho = \lbrace m_1 \mapsto a_1, m_2 \mapsto a_2 \rbrace$.
Conventions of this kind in a signaling problem are what Lewis calls \emph{signaling systems}.
An example of complete miscoordination would be $\sigma = \lbrace t_1 \mapsto m_1, t_2 \mapsto m_2 \rbrace$ and $\rho = \lbrace m_1 \mapsto a_2, m_2 \mapsto a_1 \rbrace$.
Partial coordination is achieved, for example, by $\sigma = \lbrace t_1 \mapsto m_1, t_2 \mapsto m_1 \rbrace$ and $\rho = \lbrace m_1 \mapsto a_1, m_2 \mapsto a_2 \rbrace$.

@@

Lewis signaling games can be seen as an alternative to the classical picture, even though the author himself might not have agreed with this use of his contribution.
In order to get there, some aspects of the approach need to be refined.
Brian Skyrms~\parencite*[80--104]{skyrms_evolution_1996} identifies some problems with the story so far.
Lewis' account of the stability of conventions rests on what could be considered strong demands for there to be a certain degree of required common knowledge between the players.
Namely, there needs to be a state of affairs that indicates to everyone involved that a certain regularity will hold, as well as ``mutual ascription of some common inductive standards and background information, rationality, mutual ascription of rationality, and so on''~\parencite*[56--57]{lewis_convention_1969}.
These requirements can seem excessive, even more so if we consider how simple signaling systems are when compared to human languages.
The models were introduced in order to help explain how language could get off the ground as a conventional system without any sort of prior agreement.
However, if we consider the origins of language from a historical perspective, it seems implausible to assume a high degree of rationality of the agents that started making use of primordial signaling systems which (hypothetically) evolved into languages.
Furthermore, communication through simple message exchange is something that almost all animals do: monkeys use calls, birds use singing, bees use dances, ants use pheromone trails, and so on.
A plausible account of the origin of language should first explain how signaling systems like those could get started, without assuming a great deal of rationality from the part of the agents involved.

@@

Skyrms' evolutionary game theory approach to signaling games not only gives more plausible grounds to support Lewis' discussion of convention, but it also accomplishes an important conceptual change: it moves most of the theory and mathematical formalisms to the descriptive side of the investigation.
Utility represents how the modeler views the signaling problem and understands the relative advantages or disadvantages of different possible strategy combinations.
Dynamics describe how strategies can evolve when driven by mechanisms of utility maximization.
Focus is put on understanding how various ingredients to the model interact, and which results they produce, not on metaphysical concerns.
While the general framework manages to abstract quite some detail away from the formalization, it does leave room for them, especially when it comes to the dynamics.
We already mentioned the replicator equation that can be seen as representing biological or cultural evolution, but one can also use dynamics inspired by learning mechanisms, or even ones assuming a high degree of knowledge of the game and other players\todo{give examples and references?}.
This range of options goes hand in hand with assumptions of rationality, from nothing more than survival of the fittest in a biologically-inspired setting, to a certain degree of rationality in the case of dynamics that attempt to mimic mechanisms of iterative learning, to higher levels of rationality and even game-theoretical reasoning.
Each of these can be utilized depending on the problem that one is interested in characterizing.

>>>>

Signaling games were first introduced as models of communication by David Lewis~\parencite*{lewis_convention_1969}.
His original objective was to address arguments raised against the possibility of language having started as a conventional system: if language is a convention, it had to be originally established by an agreement; in order to establish an agreement, a convention-governed system of communication would have to already have been in place; thus, although some languages could have been established by agreement if another convention was already in place, not all of them could.
In order to support the idea that conventions can arise without any prior conventional activity, Lewis studies coordination problems formalized in terms of game theory.
These are ``situations of interdependent decision by two or more agents in which coincidence of interest predominates and in which there are two or more proper coordination equilibria''~\parencite*[24]{lewis_convention_1969}.
In game theory terms, the agents interested in the coordination are the players, the game involves each player making an independent choice from his set of available actions, a payoff is what is attributed to each player based on the choices of both.
Adapting one of Lewis' examples, say Alice and Bob want to get together.
They usually meet at either Caf\'e One or Bistro Two.
Imagine there is no way for them to make any explicit agreement about it.
They are thus left to independently decide to either go to Caf\'e One or go to Bistro Two and hope for the other to show up there.
Neither has any preference for either place, but they do want to meet.
Each thus prefers to go to one of the places only if the other also decides to go to that particular place.

In order to extend his notion of convention to \emph{linguistic} exchanges, Lewis considers situations where the actions available involve sending and receiving signals or messages.
Thus, we could think of two players with different roles.
The first player, the sender, has knowledge about which of a number of possible states of affairs obtains and, depending on this information, chooses a signal to send.
The second player, the receiver, has knowledge about which signal the sender chose and, based on this information, chooses one of several possible responses.
A preference relation exists between responses and states of affairs, and a payoff is attributed to each player based on the choices of both.
Note that Lewis assumes that no player has any preference regarding the particular signal that is used, provided that it enables coordination.
Formally, in order to describe the setup all we need is to specify a set of possible states of affairs $T$, a set of available signals or messages $M$, a set of responses or actions $A$, and the utility function $U : T \times A \rightarrow \mathbb{R}^2$.

Despite the added dimension of the signal exchange, these so-called signaling problems can be seen as particular cases of coordination problems if we consider the players' choices to be of contingency plans or strategies.
A communicator's contingency plan, or sender strategy as we will call it, is a specification of a choice of message for each possible state of affairs.
It thus describes the sender's behavior conditional on the state of affairs that obtains.
An audience's contingency plan, or receiver strategy as we will call it, analogously specifies a choice of action for each possible message.
Thus, formally, what the sender chooses is a function $\sigma : T \rightarrow M$ and the receiver a function $\rho : M \rightarrow A$.
The expected utility $EU$ of a pair of strategies $(\sigma,\rho)$ can be calculated using the utility function between states of affairs and actions as a sum of the payoffs for all cases. %, \emph{i.e.}:
% $$
% EU(\sigma, \rho) = \sum_{t \in T} U(t, \rho(\sigma(t)))
% $$
As an example, consider a game with $T = \lbrace t_1, t_2 \rbrace$, $M = \lbrace m_1, m_2 \rbrace$, $A = \lbrace a_1, a_2 \rbrace$, and the following utility matrix:
\begin{center}
\begin{tabular}{r|c|c|}
\multicolumn{1}{r}{}
 & \multicolumn{1}{c}{$a_1$}
 & \multicolumn{1}{c}{$a_2$} \\ \cline{2-3}
   $t_1$ & $1,1$ & $0,0$ \\ \cline{2-3}
   $t_2$ & $0,0$ & $1,1$ \\ \cline{2-3}
\end{tabular}
\end{center}
Possible sender and receiver strategies are, for example, $\sigma = \lbrace t_1 \mapsto m_2, t_2 \mapsto m_1 \rbrace$ and $\rho = \lbrace m_1 \mapsto a_2, m_2 \mapsto a_1 \rbrace$.
These would have an expected utility of $2$ for both sender and receiver, since when $t_1$ obtains the sender will use $m_2$ and to this message the receiver will respond with $a_1$ which achieves a payoff of $1$, when $t_2$ obtains the sender will use $m_1$ and to this message the receiver will respond with $a_2$ which also achieves a payoff of $1$.
They also represent one of the two stable conventions in this game, the other being the pair of strategies $\sigma = \lbrace t_1 \mapsto m_1, t_2 \mapsto m_2 \rbrace$ and $\rho = \lbrace m_1 \mapsto a_1, m_2 \mapsto a_2 \rbrace$.
Conventions of this kind in a signaling problem are what Lewis calls \emph{signaling systems}.
An example of complete miscoordination would be $\sigma = \lbrace t_1 \mapsto m_1, t_2 \mapsto m_2 \rbrace$ and $\rho = \lbrace m_1 \mapsto a_2, m_2 \mapsto a_1 \rbrace$.
Partial coordination is achieved, for example, by $\sigma = \lbrace t_1 \mapsto m_1, t_2 \mapsto m_1 \rbrace$ and $\rho = \lbrace m_1 \mapsto a_1, m_2 \mapsto a_2 \rbrace$.

@@

Lewis signaling games can be seen as an alternative to the classical picture, even though the author himself might not have agreed with this.
In order to get there, some aspects of the approach need to be refined.
Brian Skyrms~\parencite*[80--104]{skyrms_evolution_1996} identifies some problems with the story so far.
Lewis' account of the stability of conventions rests on what could be considered strong demands for there to be a certain degree of required common knowledge between the players.
Namely, there needs to be a state of affairs that indicates to everyone involved that a certain regularity will hold, as well as ``mutual ascription of some common inductive standards and background information, rationality, mutual ascription of rationality, and so on''~\parencite*[56--57]{lewis_convention_1969}.
These requirements can seem excessive, even more so if we consider how simple signaling systems are when compared to human languages.
The models were introduced in order to help explain how language could get off the ground as a conventional system without any sort of prior agreement.
However, if we consider the origins of language from a historical perspective, it seems implausible to assume a high degree of rationality of the agents that started making use of primordial signaling systems which (hypothetically) evolved into languages.
Furthermore, communication through simple message exchange is something that almost all animals do: monkeys use calls, birds use singing, bees use dances, ants use pheromone trails, and so on.
A plausible account of the origin of language should first explain how signaling systems like those could get started, without assuming a great deal of rationality from the part of the agents involved.

@@

Skyrms' evolutionary game theory approach to signaling games not only gives more plausible grounds to support Lewis' discussion of convention, but it also accomplishes an important conceptual change: it moves most of the theory and mathematical formalisms to the descriptive side of the investigation.
Utility represents how the modeler views the signaling problem and understands the relative advantages or disadvantages of different possible strategy combinations.
Dynamics describe how strategies can evolve when driven by mechanisms of utility maximization.
Focus is put on understanding how various ingredients to the model interact, and which results they produce, not on metaphysical concerns.
While the general framework manages to abstract quite some detail away from the formalization, it does leave room for them, especially when it comes to the dynamics.
We already mentioned the replicator equation that can be seen as representing biological or cultural evolution, but one can also use dynamics inspired by learning mechanisms, or even ones assuming a high degree of knowledge of the game and other players\todo{give examples and references?}.
This range of options goes hand in hand with assumptions of rationality, from nothing more than survival of the fittest in a biologically-inspired setting, to a certain degree of rationality in the case of dynamics that attempt to mimic mechanisms of iterative learning, to higher levels of rationality and even game-theoretical reasoning.
Each of these can be utilized depending on the problem that one is interested in characterizing.
Thus, although Skyrms shows that rationality is not necessary for signaling conventions to evolve, the framework does leave room for the study of linguistic interactions where rationality can play a role.
